---
title: "Linear Regression - Matrix Method amd lm() function"
author: "Dipin P Joseph"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:

  prettydoc::html_pretty:
    
    theme: cayman
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Linear Regression using matrix method.

Defining matrix of Predictor Variable, X,
```{r}
X <- matrix(c(3,4,5),nrow = 3)
X
```

Defining matrix of Outcome Variable, Y
```{r}
Y <- matrix(c(5,9,13),nrow = 3)
Y
```

Defining X matrix with an extra column of 1's
```{r}
one_col <- rep(1,3)
X_1 <- matrix(rbind(one_col,c(3,4,5)),nrow = 3,byrow = TRUE)
X_1
```

<script async src="https://pagead2.googlesyndication.com/pagead/js/adsbygoogle.js"></script>
<!-- Notebook Check -->
<ins class="adsbygoogle"
     style="display:block"
     data-ad-client="ca-pub-3600373850128255"
     data-ad-slot="2651335722"
     data-ad-format="auto"
     data-full-width-responsive="true"></ins>
<script>
     (adsbygoogle = window.adsbygoogle || []).push({});
</script>

Calculating Beta matrix with the equation - B=((X'X)^-1)(X'Y)

```{r}
B<-solve(t(X_1)%*%(X_1))%*%t(X_1)%*%Y
B
```

Coefficients for linear regression, Slope-B1, Intercept-B0
```{r}
B0 = B[1]
B1 = B[2]
B0
B1
```

Finding residuals of the linear regression model, Res = Y-(B1*X+B0) which equals zero.
```{r}
Y_Pred <-(B1*X+B0)
Res = Y-Y_Pred
Res
```

<script async src="https://pagead2.googlesyndication.com/pagead/js/adsbygoogle.js"></script>
<!-- Notebook Check -->
<ins class="adsbygoogle"
     style="display:block"
     data-ad-client="ca-pub-3600373850128255"
     data-ad-slot="2651335722"
     data-ad-format="auto"
     data-full-width-responsive="true"></ins>
<script>
     (adsbygoogle = window.adsbygoogle || []).push({});
</script>


# Linear Regression using lm() function


```{r}
l_reg <- lm(formula = Y ~ X)
summary(l_reg)
```

From the summary, R square value is 1 and residual error 0 which implies that model is a perfect fit but highly dependent on perdictor variable and is unrealistic.


Plotting X,Y, predicted values and linear model.
```{r echo=FALSE}
# Scatter plot for X and Y
plot(main = 'Linear Regression Y~X',X,Y, xlim = c(0,6) , ylim = c(-10,15), pch=19, col="#13871c", cex=1)
# Predicted values for X with regression
Y_pred <- predict.lm(l_reg, newdata = list(x=X))
# Points with X and predicated vaues of Y
points(X, Y_pred, col="#d37024", cex=2)
# Linear regression line
abline(l_reg, col="#3f4da7", lwd=3)
legend(0,15, legend=c("Predicted Y", "Actual Y"),
       col=c("#d37024", "#13871c"), pch=c(1,19), cex=0.8)
text(x=1, y=-2, labels="y = β0 + β1X = -7+4x", col="#3f4da7",srt=27)
text(x=0.55, y=-2.75, labels="^     ^      ^", col="#3f4da7",srt=27)
```

<script async src="https://pagead2.googlesyndication.com/pagead/js/adsbygoogle.js"></script>
<!-- Notebook Check -->
<ins class="adsbygoogle"
     style="display:block"
     data-ad-client="ca-pub-3600373850128255"
     data-ad-slot="2651335722"
     data-ad-format="auto"
     data-full-width-responsive="true"></ins>
<script>
     (adsbygoogle = window.adsbygoogle || []).push({});
</script>
