---
title: "Outlier Detection in R"
author: "Dipin P Joseph"
date: "26/04/2020"
output:

  prettydoc::html_pretty:
    
    theme: cayman
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this post, we will go through different outlier detection methods and topics like,

* Power Transformation (Yeo Johnson Transform) and Density Plots
* Box Plots and 5 Point Summary Statistics
* Z-Score
* Histograms and Scatter Plots
* Mosaic Plots and Bag Plots
* Mahalanobis Distance
* Cook's Distance
* DBScan 
* Local Outlier Factors
* 1-class Support Vector Machine

Dataset under consideration - **Males** (from package - _Ecdat_)

First, we will explore more about data using __str()__, __summary()__ methods,

```{r}
library(Ecdat)
summary(Males)
str(Males)
```

## Train-Test Split

Here, we will split data to training and testing set. 70% will go for training and 30% for testing.

```{r}
set.seed(7)
# 70% of the sample size
smp_size <- floor(0.7 * nrow(Males))

# Train and Test Split
train_ind <- sample(seq_len(nrow(Males)), size = smp_size)
tr <- Males[train_ind,]
te <- Males[-train_ind,]

# Column names split
cols <- colnames(Males)
cols_num <-cols[3:4]
cols_num_all <- c(cols_num, cols[9])
cols_cat <- c(cols[5:8],cols[10:12])
```

## Yeo-Johnson Transformation

Most of the methods will only work well with normalized data. So, we have to transform our variables to another form which are more normally distributed. Yeo-Johnson tansformation will help us here,

```{r}
library(plotly)
library(recipes)

# Performing Yeo Johnson transform for Normalization
# Setting variable price as outcome and remaining numerics as predictors
yj_estimates <-
  recipe(as.formula(paste("wage ~ ", paste(cols_num, collapse = " + "))), data = tr) %>%
  # Power transformation step
  step_YeoJohnson(all_numeric()) %>%
  # Feeds training data
  prep(data = tr)
# The trained process is run of test set
yj_t <- bake(yj_estimates, te)

```

## Density Plot

To display effect of YJ transform density of one variable before and after transformation is plotted.

```{r}
de_b <- density(te$exper)
de_b <- data.frame(x = de_b$x, y = de_b$y)
before <- plot_ly(data=de_b, x = ~x, y = ~y)
layout(
  add_lines(before),
  title = "Density plot of non-transformed variable"
)
```
```{r}
de_a <- density(yj_t$exper)
de_a <- data.frame(x = de_a$x, y = de_a$y)
after <- plot_ly(data=de_a, x = ~x, y = ~y)
layout(
  add_lines(after),
  title = "Density plot of transformed variable"
)
```
## Univariate Outlier Detection

## Box Plots

Now, let's plot box plots for numeric variables in the dataset. If you see a datapoint outside the whiskers, then you need to examine that particular observation.

```{r}
        bp <- plot_ly(data = yj_t, type = 'box')
        for (k in 1:length(cols_num_all)) {
            dfk <- data.frame(y = yj_t[[cols_num_all[k]]])
            bp <-
                add_trace(
                    bp,
                    y = ~ y,
                    data = dfk,
                    name = cols_num_all[k],
                    notched = TRUE,
                    text =  ~ y
                )
        }
        
        layout(bp,
               yaxis = list(title = "Value"))
```

## Five Number Summary

Boxplots are created on basis of 5 number summary. This includes,

* Minimum
* Q1 - 25%
* Median
* Q3 - 75%
* Maximum

Let's detect 5 number summary for variable __wages__,

```{r}
  fnum = (fivenum(yj_t$wage))
  print(fnum)
  # Calculating end point of whiskers
  low = fnum[2] - 1.5 * (fnum[4] - fnum[2])
  high = fnum[4] + 1.5 * (fnum[4] - fnum[2])
  print(paste("Outliers are outside region",low,high))
```

